Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-10 review_article.Rmd
done revising the choropleth map section, material on atlases shifted to the next section

<!-- Structure of the paper -->
The paper is organised as follows.  The next section, [Map displays for disease data](#choropleths), describes the choropleth map which is the common approach to disease maps. Section [Publicly available atlases](#public) surveys atlases in use today. The [Alternative displays](#alternative) describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section [Comparison of mapping methods](#compare). Disease maps are more useful when made interactive, and common options are described in Section [Animation and Interactivity](#interacting), along with a discussion of benefits and disadvantages. The last section summarises the paper and discusses future directions.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the characteristics of a spatial relationship of measurements. 
They are true maps of the topology, constructed for visual inspection of spatial patterns on a familiar geographic form. Classed choropleth maps break a continuous variable into a small set of categories, often quintiles, and use a gradual colour scheme to shade all areas that fall in the same category [@CPISACA].
Figure \ref{fig:choro} shows a classed choropleth map of the United States. Each state has been coloured according to the average age-adjusted rate per 100,000 people, of new cases of lung and bronchus for females and males from 2012 to 2016.
The data was extracted from the [official federal statistics](https://gis.cdc.gov/cancer/USCS/DataViz.html) on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI).

Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

@@

b <- st_bbox(cancer)
ggplot(cancer) + 
  geom_sf(aes(fill = AgeAdjustedRate)) +
  scale_fill_distiller("Age-adjusted rate (per 1000)", type = "seq", palette = "YlOrRd",  direction = 1) + 
  #coord_sf(crs = CRS("+init=epsg:3857"), xlim = c(b["xmin"], b["xmax"]), ylim = c(b["ymin"], b["ymax"])) +
  theme_void() +
  theme(legend.position ="bottom", legend.key.width = unit(2, "cm"))
```

<!-- psychology -->
Utilising the familiar state boundaries can make a map intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data, i.e. how cancer rate differs across the states.
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.
While the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the unpopulated but large areas [@EI]. 
Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

@@

# Publicly available atlases

@@

## Overview

@@

## Supplementary statistics in public facing atlases

@@

# Alternative displays {#alternative}

@@

### Alternatives

@@

@@

# Comparison of mapping methods (#compare)

>>>>

<!-- Structure of the paper -->
The paper is organised as follows.  The next section describes the choropleth map which is the common approach to disease maps. Section \ref{public} surveys atlases in use today. The (#alternative) describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section [Comparison of mapping methods](#compare). Disease maps are more useful when made interactive, and common options are described in Section [Animation and Interactivity](#interacting), along with a discussion of benefits and disadvantages. The last section summarises the paper and discusses future directions.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the spatial relationship of measurements. The geography  is faithfully rendered, and the colour rendering is designed to reveal spatial patterns among data values. A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI]. 
Figure \ref{fig:choro} shows a choropleth of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the [official federal statistics](https://gis.cdc.gov/cancer/USCS/DataViz.html) on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI).

<!-- history and types of choropleths, variations -->
Early versions of choropleth maps used symbols or patterns instead of colour.
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO] XXX Is this the right reference here? It doesn't seem to match the statement of early examples. Maybe you want to reference Dupin's map? @CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

<!-- psychology -->
Utilising the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data, i.e. how cancer rate differs across the states. The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. In Figure \ref{fig:choro}, a west to east spatial trend  of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbours. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbours. There is something of a cluster of higher rates around the tobacco states.

@@

b <- st_bbox(cancer)
ggplot(cancer) + 
  geom_sf(aes(fill = AgeAdjustedRate)) +
  scale_fill_distiller("Age-adjusted rate \n (per 100,000)", type = "seq", palette = "YlOrRd",  direction = 1) + 
  #coord_sf(crs = CRS("+init=epsg:3857"), xlim = c(b["xmin"], b["xmax"]), ylim = c(b["ymin"], b["ymax"])) +
  theme_void() +
  theme(legend.position ="bottom", legend.key.width = unit(2, "cm"))
```

<!-- potential problems -->
While the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the unpopulated but large areas [@EI]. 
Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

# Cancer atlases

@@

@@

## Overview of publicly available atlases

@@

XXX more details on the examples in the next figure

## Common statistics displayed

@@

## Interaction with atlases

# Cartograms and other spatially warped displays {#alternative}

@@

### Dorling

@@

## Tilegram

@@

Bob Rudis, geogrid

# Comparison of maps and cartograms {#compare}
